The shape of kelp: environmental influences on kelp morphology

### 2018-11-08 chapter_2.Rmd
some minor additions to discussion
added to background section and begining of conclusion

Kelps are sessile species and have to adapt to local environmental conditions in order to persist and survive. The resilience of kelps is evident through their distribution, which spans a range of environmental conditions that vary spatially and temporally. Wave exposure and temperature are regarded as important environmental drivers with regards to macroalgae, and play a role in the distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of kelps. Wave exposure is recognised as the main cause of kelp mortality, which may manifest itself through dislodgment of individuals or breakage of important morphological structures. These mechanisms usually occur during times of high wave energy or pulse disturbance events, such as storms.Mechanical forces generated by the hydrodynamic environment, in the form of sudden strong ocean currents or storms, between 10- 20 m s-1 with accelerations of 400 m s-2 [@Friedland1995] are the biggest threat to kelp survival. 
Kelps are able to reduce the probability of dislodgment or breakage through morphological adaptation, which may involve several strategies. Kelps can either reduce drag, increase attachment strength or increase flexibility in order to survive high wave energy environments.The strategy that the algal organism employs is dependent on the species. For instance, branched, articulated algae will vary its morphology differently to that of foliose, more flexible macroalgae. For example, @moss1948 investigated the anatomy, chemical composition of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation physiological components, such as organic nitrogen, mannitol, laminarian and alginic acid concentrations. The authors also noted a ‘crumpling effect’ displayed by individuals from exposed sites and inferred that this strategy may reduce overall drag.  Other studies show that macroalgae in wave exposed environments have morphologies that reduce overall drag, increase strength of attachment or increase flexibility. For example, a study by @FowlerWalker2006 tested for differences in morphology of *Ecklonia radiata* between wave-sheltered and wave-exposed sites and through a combination of *in situ* sampling and transplantation of juvenile plants. The results showed that morphology differed between wave-sheltered and wave-exposed sites (thin thallus at sheltered sites and a narrow, thick thallus with a thick stipe at exposed sites), and was consistent with previous studies. Juveniles transplanted into wave exposed sites under went rapid morphological adaptation, whilst the opposite was true for wave-sheltered sites which showed slower morphological adaptation. Dislodgement through increased wave exposure is not the only hydrological parameter playing a role in kelp mortality. A study by @DeBettignies2013 tested if there is a linked between orbital water velocity representative of wave-swept reefs and the morphology and biomass of *Ecklonia radiata*. The results showed that morphological adaptation through drag reduction and strength increase strategies do not reduce drag as efficiently as reduction in overall biomass through reduction in thallus surface area. The authors concluded that at peak orbitat water velocities size (biomass) and not morphological adaptation was the main driver of hydrological performance in wave-swept environments. There is also evidence that morphological adaptation is driven by currents, and in fact may be driving hydrological performance of macrolgae.

@@

Seasonal variations in significant wave height (Hs), wave period and wave direction were observed from the data. The direction of swell swings to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006]. These swells were found to correlate with increased wave period (Tp), wave direction and Hs. False Bay is therefore shielded by these south westerly swells by the Cape Peninsula [@Shipley1964]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an explanation for increased variability of Hs and Tp. The wind speed and wind direction support the presence of upwelling in summer, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a]. This is supported by the decreased in temperature in summer, specifically along the west side of Cape Point. We however see a differentiation of median wind directions in summer, with the western coast of False Bay experiencing south easterly winds, compared to the west side of Cape Point experiencing south westerly winds. It is hypothesised that the topography and elevation along the Cape Peninsula channels, shields winds along the strip of land. This is however absent in winter, where strong northerly winds are experiences from St. Helena Bay to Betty’s Bay [@Field1980a]. St. Helena bay experiences decreased median wave directions, and is protected by a headland, similar to the sites found along the western side of False Bay, that are shielded by the Cape Peninsula. There is however increased variability in the wave direction from both False Bay sites and St. Helena Bay, which is encouraged by refraction of waves.

@@

>>>>

Kelps are sessile species and have to adapt to local environmental conditions in order to persist and survive. The resilience of kelps is evident through their distribution, which spans a range of environmental conditions that vary spatially and temporally. Wave exposure and temperature are regarded as important environmental drivers with regards to macroalgae, and play a role in the distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of kelps. Wave exposure is recognised as the main cause of kelp mortality, which may manifest itself through dislodgment of individuals or breakage of important morphological structures. These mechanisms usually occur during times of high wave energy or pulse disturbance events, such as storms.Mechanical forces generated by the hydrodynamic environment, in the form of sudden strong ocean currents or storms, between 10- 20 m s-1 with accelerations of 400 m s-2 [@Friedland1995] are the biggest threat to kelp survival. 
Kelps are able to reduce the probability of dislodgment or breakage through morphological adaptation, which may involve several strategies. Kelps can either reduce drag, increase attachment strength or increase flexibility in order to survive high wave energy environments.The strategy that the algal organism employs is dependent on the species. For instance, branched, articulated algae will vary its morphology differently to that of foliose, more flexible macroalgae. For example, @moss1948 investigated the anatomy, chemical composition of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation physiological components, such as organic nitrogen, mannitol, laminarian and alginic acid concentrations. The authors also noted a ‘crumpling effect’ displayed by individuals from exposed sites and inferred that this strategy may reduce overall drag.  Other studies show that macroalgae in wave exposed environments have morphologies that reduce overall drag, increase strength of attachment or increase flexibility. For example, a study by @FowlerWalker2006 tested for differences in morphology of *Ecklonia radiata* between wave-sheltered and wave-exposed sites and through a combination of *in situ* sampling and transplantation of juvenile plants. The results showed that morphology differed between wave-sheltered and wave-exposed sites (thin thallus at sheltered sites and a narrow, thick thallus with a thick stipe at exposed sites), and was consistent with previous studies. Juveniles transplanted into wave exposed sites under went rapid morphological adaptation, whilst the opposite was true for wave-sheltered sites which showed slower morphological adaptation. Dislodgement through increased wave exposure is not the only hydrological parameter playing a role in kelp mortality. A study by @DeBettignies2013 tested if there is a linked between orbital water velocity representative of wave-swept reefs and the morphology and biomass of *Ecklonia radiata*. The results showed that morphological adaptation through drag reduction and strength increase strategies do not reduce drag as efficiently as reduction in overall biomass through reduction in thallus surface area. The authors concluded that at peak orbitat water velocities size (biomass) and not morphological adaptation was the main driver of hydrological performance in wave-swept environments.

There is also evidence that morphological adaptation is driven by currents, and in fact may be driving hydrological performance of macrolgae. @Duggins2003 examined the direct and indirect flow effects on population dynamics, morphology and biomechanics of several understorey macroalgae species. These species included *Costaria costata*, *Agarum fimbriatum*, and *Laminaria complanata* and * Nereocystis luetkeana*. The results showed that in wave impacted sites (wave exposed) had higher rates of mortality, and no significance was found between survival of individuals and tidal or current velocity. The authors concluded that although tidal and current velocity did not play a significant role in determining kelp survival, it did play a role in morphological adaptation. The results from this study suggest that high current and tidal stresses are the main driver of kelp morphological adaptation. This in turn make those individuals more resilient to dislodgement to wave exposure. Wave exposure is stochastic in nature compared to tidal and ocean currents which are more regular in their frequency and magnitude. Therefore the regular forces of tidal and ocean currents may make kelp individuals more resilient to mechanical dislodgement over time.

@@

Seasonal variations in significant wave height (Hs), wave period and wave direction were observed from the data. The direction of swell swings to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006]. These swells were found to correlate with increased wave period (Tp), wave direction and Hs. False Bay is therefore shielded by these south westerly swells by the Cape Peninsula [@Shipley1964]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an explanation for increased variability of Hs and Tp. However, changing winds may also be influencing waves generated within False Bay. As seen in Figure 1 the summer and winter swells are relativley constant in terms of direction adn this may be driving the waves locally (I am still looking for the report where Christo cites this). The wind speed and wind direction support the presence of upwelling in summer, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a]. This is supported by the decreased in temperature in summer, specifically along the west side of Cape Point. We however see a differentiation of median wind directions in summer, with the western coast of False Bay experiencing south easterly winds, compared to the west side of Cape Point experiencing south westerly winds. It is hypothesised that the topography and elevation along the Cape Peninsula channels, shields winds along the strip of land. This is however absent in winter, where strong northerly winds are experiences from St. Helena Bay to Betty’s Bay [@Field1980a]. St. Helena bay experiences decreased median wave directions, and is protected by a headland, similar to the sites found along the western side of False Bay, that are shielded by the Cape Peninsula. There is however increased variability in the wave direction from both False Bay sites and St. Helena Bay, which is encouraged by refraction of waves.

@@

## Conclusion

(More to add here but I wanted to add it so long)
Mechanical dislodgment through hydrological stress is the main source on mortality for kelp, and previous research has shown that wave exposure is the main driver of kelp morphology. Therefore, morphological adaptation is the main strategy to reduce drag and ultimately dislodgement in hydrological stressful environments. The results from this study support previous findings, and shows that particular wave parameters (significant wave height and median wave direction) act more strongly in driving kelp morphological plasticity.
